# Personal movement activity

```{r message=FALSE}
library(stringr)
library(lubridate)
library(dplyr)
library(readr)
```

### 1. Code for reading in the dataset and/or processing the data

```{r message=FALSE}
df <- read.csv("activity.csv")
head(df)
```

A closer look at the <tt>interval</tt> variable reveals that the intervals represent the time of the day in hours and minutes, with omitted leading zeros: in the first row of the data frame we thus have the number of steps taken on October 1, 2012, between 00:00 and 00:05 (i.e. between midnight and five minutes past midnight).

```{r}
df[1000:1010,]
```

From the data frame snippet above we see, for example, that on October 4, 2012, the subject took 180 steps between 11:25 and 11:30. We can now proceed to aggregate the data and compute the total number of steps taken in each day, and use the outcome to create a histogram. 

```{r}
df_agg <- aggregate(df$steps, list(date=df$date), sum)
names(df_agg)[names(df_agg)=="x"] <- "steps"
head(df_agg)
```

### 2. Histogram of the total number of steps taken each day
The difference between histogram and bar plot is important here. We should bear in mind that our goal is not to plot the date on the x-axis and the total number of steps on the y-axis, but rather to create bins of number of steps (whereby the size of the bins is arbitrary) in order to represent the frequency (count) of a certain range of steps (e.g., how many days did the subject take a number of steps that lies between 4000 and 5000?).  

```{r Plot_1, fig.cap = "Plot 1"}
par(bg = 'gray95')
with(df_agg, hist(steps, breaks=20, col="lightblue", xlim = c(0, 25000),
                  main="Number of steps"))
```

The histogram tells us, for example, that there are 10 days in which the number of steps taken was between 10000 and 11000, and 1 day in which that number was between 4000 and 5000. It is important to point out that NA values are ignored, therefore the days in which all the intervals have value NA are not displayed (8 days are missing from the total count).

### 3. Mean and median number of steps taken each day
```{r}
mean_steps <- round(mean(df_agg$steps, na.rm=TRUE), digits=0)
print(paste("Mean of steps taken each day (rounded to integer):", mean_steps))

median_steps <- median(df_agg$steps, na.rm=TRUE)
print(paste("Median of steps taken each day:", median_steps))

total_steps <- sum(df_agg$steps, na.rm=TRUE)
print(paste("Total number of steps:", total_steps))
```

### 4. Time series plot of the average number of steps taken
The aim of this plot is to show all the 5-minute intervals in one day on the x-axis, and on the y-axis the average of steps taken in each given interval (the average is computed over the entire two-month period). We want to be able to answer questions such as: "What's the average of steps taken in the interval between 9:25 and 9:30, over the two months?"  
The number of intervals in one day is 12 x 24 = 288 (there are 12 intervals in one hour and 24 hours in a day). In order to enable the conversion to a datetime object, we pad the interval with leading zeros (to give the format HHMM) and with trailing zeros (which results in the format HHMMSS).

```{r}
df$interval <- str_pad(df$interval, width=4, side="left", pad="0")
df$interval <- str_pad(df$interval, width=6, side="right", pad="0")
head(df)
```

We then aggregate the data to obtain the average number of steps taken in each interval, over the two months, and we convert the interval into the format HH:MM:SS

```{r}
df_agg_int <- aggregate(df$steps, list(interval=df$interval), mean, na.rm=TRUE)
df_agg_int$interval <- parse_time(df_agg_int$interval, "%H%M%S")
names(df_agg_int)[names(df_agg_int)=="x"] <- "avg.steps"
df_agg_int$avg.steps <- round(df_agg_int$avg.steps, digits=2)
tail(df_agg_int)
```

We can see, for example, that in the interval betwenn 23:30 and 23:35, the average number of steps taken is 2.60. Now we create a time series. In order to enable the creation of the plot, we need a dummy date (in this case "2001-01-01"), though this date has no actual meaning.

```{r}
df_agg_int <- mutate(df_agg_int, time_series = ymd_hms(paste("2001-01-01", interval)))
```

```{r Plot_2, fig.cap = "Plot 2"}
par(bg = 'gray95')
with(df_agg_int, plot(time_series, avg.steps, type = 'l',
                      xlab = "interval", ylab = "average steps", ylim=c(0, 220),
                      main = "Average steps across intervals"))
```

The graph shows that, on average, there is a peak in the number of steps taken somewhere between 08:00 and 09:00, whereas the average is very close to zero in the intervals between midnight and 06:00.
The next chunk, which is not evaluated, can be used as an alternative to the previous one, in case we want to plot the difference from the mean instead of the value. The shape of the plot will be the same, only the scale on the y-axis will change.

```{r eval=FALSE}
# The next two lines will compute the difference between the average number of 
# steps for one inerval and the mean number of steps across all intervals.
g <- mean(df_agg_int$avg.steps)
df_agg_int <- mutate(df_agg_int, diff = avg.steps - g)

with(df_agg_int, plot(time_series, diff, type = 'l',
                      xlab = "interval", ylab = "average steps"))
```


### 5. The 5-minute interval that, on average, contains the maximum number of steps
We can find out more precisely which 5-minute interval has the maximum average of steps taken.

```{r}
max_steps <- df_agg_int[which.max(df_agg_int$avg.steps), c(1,2)]
print(max_steps)
```

### 6. Code to describe and show a strategy for imputing missing data

Let's take a look at the original data frame:

```{r}
parsed.interval <- parse_time(df$interval, "%H%M%S")
df <- cbind(df, parsed.interval)
head(df)
```

We observe missing values for some of the intervals, and we can find out that their number is 2304:

```{r}
print(paste("Total number of entries (intervals):", length(df$steps)))
table(is.na(df$steps))
```

Similarly, in the aggregated data frame we can find out that there are 8 days in which data are missing for all the intervals (one example is 2012-10-01):

```{r}
head(df_agg)
print(paste("Total number of entries (days):", length(df_agg$steps)))
table(is.na(df_agg$steps))
```

One possible strategy to impute missing data could be to replace all the missing intervals with the mean number of steps for that particular interval over the two months.

```{r}

# Initialize counters
i <- 1
j <- 1

# Create empty vectors which will be variables in 
# a data frame used to record imputed values
ID <- vector()
ID_NA <- vector()
date <- vector()
interval <- vector()
imputed_value <- vector()

# Create an empty vector for the
# variable that will replace df$steps
new_steps <- vector()

for (item in df$steps) {
  if (is.na(item)) {
    k <- df_agg_int$avg.steps[match(df$parsed.interval[i], df_agg_int$interval)]
    ID <- c(ID, i)
    ID_NA <- c(ID_NA, j)
    date <- c(date, as.character(df$date[i]))
    interval <- c(interval, as.character(df$parsed.interval[i]))
    imputed_value <- c(imputed_value, k)
    j <- j+1
    # If value is NA, add the imputed value
    new_steps <- c(new_steps, k)
  }
  # If value is not NA, keep the existing value
  else {new_steps <- c(new_steps, item)}
  i <- i+1
}
```

Just as a sanity check, with the output of the code above we can create a data frame with all the imputed values, where ID indicates the row number in the data frame with all the intervals and ID_NA indicates the row number in the data frame that subsets only the NA values (which has 2304 entries).

```{r}
df_impute <- data.frame(ID, ID_NA, date, interval, imputed_value)
dim(df_impute)
tail(df_impute)
```

We can see, for example, that the missing number of steps for the interval from 23:30:00 to 23:35:00 has been imputed with the mean value of steps for the same interval in the entire period of two months. I have decided not to round the imputed values to be integers.
A vector <tt>new_steps</tt> has been created, with the variable <tt>df$steps</tt> now updated with the imputed value.

```{r}
length(new_steps)
head(new_steps)
df_new <- cbind(df, new_steps)
head(df_new)
```

We can now proceed to aggregate the new data, visualize the histogram with imputed values and compare it to the one in section 2 (with missing values):

```{r}
df_agg_new <- aggregate(df_new$new_steps, list(date=df_new$date), sum)
names(df_agg_new)[names(df_agg_new)=="x"] <- "steps"
df_agg_new$steps <- round(df_agg_new$steps, digits=0)
head(df_agg_new)
```

```{r Plot_3, fig.cap = "Plot 3", fig.width=12, fig.height=6}
par(bg='gray95', mfrow=c(1, 2))
with(df_agg_new, hist(steps, breaks=20, col="lightyellow",
                      xlim=c(0, 25000), ylim=c(0, 20),
                      main="Number of steps\nwith imputed values")) 
with(df_agg, hist(steps, breaks=20, col="lightblue",
                  xlim = c(0, 25000), ylim=c(0, 20),
                  main="Number of steps\nwith missing values"))
```

We observe that the count for the range between 10000 and 11000 has increased from 10 to 18. We can interpret this by stating the following: after imputing the data, the 8 days where all the intervals were NA fall in the range 10000-11000 steps. A closer look at the data frame reveals, unsurprisingly, that they actually got a number of steps = 10766, which is the mean value computed in Section 3.

```{r}
mean_steps_new <- round(mean(df_agg_new$steps), digits=0)
print(paste("Mean of steps taken each day (rounded to integer):", mean_steps_new))

median_steps_new <- median(df_agg_new$steps)
print(paste("Median of steps taken each day:", median_steps_new))

total_steps_new <- sum(df_agg_new$steps, na.rm=TRUE)
print(paste("Total number of steps:", total_steps_new))
```

After we impute the missing data, there is no change in the mean, and the median increased only by 1. The total number of steps increased from 570608 to 656736. The difference (86128) matches the product of the mean (10766) and the number of missing days (8).
We would have obtained a different result if the NA values had been converted to 0 instead of being ignored. In that case, the 8 missing days would have 0 steps, wich would result in lower values of mean and median.

### 8. Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

```{r}
df_new$date <- as.Date(df_new$date, format="%Y-%m-%d")
df_new$days <- ifelse(weekdays(df_new$date) %in%
                        c("Saturday", "Sunday"), "weekend", "weekday")
head(df_new)
```

```{r}
df_days <- aggregate(df_new$new_steps,
                     list(day=df_new$days, interval=df_new$interval), mean)
names(df_days)[names(df_days)=="x"] <- "avg.steps"
df_days$avg.steps <- round(df_days$avg.steps, digits=2)
head(df_days)
dim(df_days)
```

We now have 576 rows, because for each of the 288 intervals there are two observations (one for weekdays and one for weekends).

```{r}
df_days <- mutate(df_days, time_series = ymd_hms(paste("2001-01-01", interval)))
```

```{r Plot_4, fig.cap = "Plot 4", fig.width=12, fig.height=6}
par(bg='gray95', mfrow=c(1, 2))
with(subset(df_days, day=="weekday"), plot(time_series, avg.steps, type = 'l',
                      xlab = "interval", ylab = "average steps",
                      ylim=c(0, 250), main="Weekdays"))
with(subset(df_days, day=="weekend"), plot(time_series, avg.steps, type = 'l',
                      xlab = "interval", ylab = "average steps",
                      ylim=c(0, 250), main="Weekends"))
```

We observe an interesting pattern: during weekdays there is a significant peak at around 08:30 and a second smaller one at about 19:00 (we may surmise that the subject goes for a walk before and after work), whereas in the weekdends then activity is more evenly distributed throughout the day.



